\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-02 chapters/bttda.tex
New patient table
Comments Marc

\subsection{Higher Order Discriminant Analysis}
Higher Order Discriminant Analysis (HODA)~\cite{Phan2010} is a
supervised tensor-based feature extraction technique.
For a set of $N$ tensors of order $K$
$\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
		D_K}\right\}_n^N$, HODA finds projection matrices $\mat{U_k}$ for each mode $k$
that project a given $\ten{X}$ to a latent tensor
$\ten{G}\in\mathbb{R}^{r_1\times r_2\times\cdots\times r_K}$, usually with lower
dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
tensor-matrix mode products
\begin{equation}
	\ten{G}  = \ten{X}\mmpr{\mat{U}}
	\label{eq:HODA-backward}
\end{equation}
visualized in Figure~\ref{fig:HODA-backward}.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_backward.tikz.tex}
	\caption{A visualization of the multilinear projection obtained by Higher Order
    Discriminant Analysis (HODA) applied to a third-order tensor
    sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
		HODA finds projection matrices $\mat{U}_k$ such that maximal
		discriminability between classes is achieved in the projected latent tensors
		$\ten{G}$ with reduced dimensionality $(r_1,r_2,r_3)$.}
	\label{fig:HODA-backward}
\end{figure}
Analogous to the HOSVD, HODA is a dimensionality
reduction decomposition that results in a dense latent tensor $\ten{G}$, and
imposes an orthogonality constraint on $\mat{U}_k$ to ensure uniqueness.
However, while for the HOSVD decomposition the projection matrices
are chosen to minimize the reconstruction error, the projection matrices
$\mat{U}_k$ of HODA are optimized for maximal discriminability of tensors
$\ten{G}(n)$ belonging to classes with labels $c_n$.
This is a desirable property in a classification setting where samples are
high-dimensional tensors.

HODA optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent
tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
The goal is now to find the optimal projection matrices\todo{You should say
somewhere that the shape (r1,r2,...rK) is fixed a priori, no?}
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which can be solved by the iterative algorithm in Algorithm~\ref{alg:HODA}.
\begin{algorithm}
	\caption{Higher order Discriminant Analysis (HODA) backward solution}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per mode Singular Value Decomposition (SVD),
or as the partial HOSVD of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
%and solving the generalized eigenvalue problem
%\begin{equation}
%  = \mat{\lambda}\mat{S}_{-k,\text{w}}\mat{U}_k
%\end{equation}
The iterative process halts after a fixed amount or iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.

@@

The HODA projection in Equation~\ref{eq:hoda-backward}\todo{missing} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model is given by
\begin{equation}
	\ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$
and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
	\caption{The forward projection for HODA. By calculating activation
		patterns $A_k$, the original tensor $\ten{X}$ can approximately be
		reconstructed from projected latent tensor $\ten{G}$. $A_k$ are chosen such
		that the variability captured in the latent tensor is maximally expressed in
		the reconstructed tensor and not in the error term.}
	\label{fig:HODA-forward}
\end{figure}
\todo{not referred to in text}
\todo{I would prefer a figure in which the Ai and G are on the left and
$\hat{X}$on the right. So, rather than using a "almost equal to" sign you could
use a standard equal signal. Finally, in the caption, you could clearly define
the original(X) and reconstructed ($\hat{X}$) tensors as well as the error term.}
\todo{explicitly state why choosing this versus full model or HOPLS}

A forward model should make sure that reconstruction error is minimized and
variation captured in the latent tensor is maximally captured by the forward
projection term $\ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
\todo{not clear ("error is minimized" in the same sentence)}
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares (ALS) algorithm~\cite{Bentbib2022},%~\cite{Comon2009},
iteratively fixing all but one of the activation patterns such that
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed amount of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in Algorithm~\ref{alg:HODA-fw}.
\begin{algorithm}
	\caption{Higher order Discriminant Analysis (HODA) forward solution}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}
\todo{Reorder figures and add error term}

\subsection{Block-Term Tensor Discriminant Analysis}
After defining the forward model, we can construct our proposed block-term
tensor model.
Assuming the latent tensors $\ten{G}$
obtained by the backward projection of HODA do not achieve perfect
class separation, the error term $\ten{\mathbfcal{E}}$ in
Equation~\ref{eq:HODA-forward} should still contain some discriminative
information, which can be exploited to improve classifier
performance.
\todo{eps1 = X - G1 x {A1} still contains discriminative info, so
it is further projected onto another core tensor yielding
eps2 = eps1 - G2 x
{A2}
etc. and finally you give equation (15)
}
We thus extend the HODA feature extraction scheme with backward and
forward models defined in respectively Equations~\ref{eq:HODA-backward}
and~\ref{eq:HODA-forward} to Block-Term Tensor Discriminant Analysis
(BTTDA).
BTTDA finds multiple discriminative blocks, such that its forward
model adheres to the block-term tensor structure:
\begin{align}
	\ten{X} & = \sum_b^B\ten{G}^{(b)}\mmpr{\mat{A}^{(b)}} + \ten{\mathbfcal{E}}
	\label{eq:BTTDA-forward}
\end{align}
for $B$ extracted latent tensors $\ten{G}^{(b)}$ and residual error term
$\ten{\mathbfcal{E}}$.
Figure~\ref{fig:BTTDA} further illustrates the BTTDA model.
\begin{figure*}[t]
	\centering
	\input{figures/bttda/BTTDA.tikz.tex}
	\caption{A forward model for Block-Term Tensor Discriminant Analysis
		(BTTDA). BTTDA can extract more features
		than HODA by iteratively finding a latent tensor $\ten{G}^{(b)}$ in a
		deflation scheme.
		The HODA backward projection is first applied. Next, the
		input data is reconstructed via the HODA forward model and the
		difference between the two is found.
		Finally, this process is repeated with this difference as input data, until a
		desired number of blocks $B$ has been found.}
	\label{fig:BTTDA}
\end{figure*}
\todo{Explicitly mention that if restricted to rank one, it is parafac, and to
one block, it is hoda}

Since BTTDA is specified as a forward model, a backward modelling
procedure is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
BTTDA to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in Algorithm~\ref{alg:BTTDA}.
\begin{algorithm}
	\caption{Block-term Tensor Discriminant Analysis (BTTDA)}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_BTTDA.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \todo{ref equation}from a residual error term
$\ten{\mathbfcal{E}}^{(b)}$
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward HODA
projection
\begin{equation}
	\ten{\mathbfcal{E}}^{(b+1)} = \ten{\mathbfcal{E}}^{(b)} - \ten{G}^{(b)}
	\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(1)}=\ten{X}$.

@@

Combined, this results in hyperparameters $B$ for the number of blocks, the
threshold value for feature retention, and the ranks of the individual blocks.
While we can reasonably set the feature selection $p$-value threshold to $0.05$,
$B$ and the block ranks must be tuned to select a performant feature extraction
model.
While these hyperparameters can be set through cross-validation, this can be
computationally expensive.
To reduce the computational cost of model selection,
Algorithm~\ref{alg:model-selection} proposes a heuristic model selection
algorithm that leverages cross-validation in a greedy way per block, to
iteratively find the optimal rank for the next block given the ranks of the
previous block.
The area under the receiver operating characteristic curve (ROC-AUC) for
classification of extracted feature vectors after feature selection is used as
cross-validation score.
\begin{algorithm}
	\caption{Greedy model selection}
	\label{alg:model-selection}
	\input{chapters/bttda/alg_model_selection.tex}
\end{algorithm}
Finally, the series of blocks can be truncated to the point with
highest validation score to determine $B$.

@@

\subsection{Block contribution}
\todo{Discuss number of features}
To analyze the contribution of extra feature blocks extracted by {BTTDA} over
the first one found by HODA we pick n ERP ($K=2$) dataset
(BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
We report within-session ROC-AUC scores for training, validation and test data as a function
of the number of blocks increases, Figure~\ref{fig:blocks}.
Training and validation folds were taken from the model selection procedure.
Additionally, the Normalized Mean Squared Error (NMSE) is reported for the
reconstructed from the truncated BTTDA decomposition
$\textstyle{\ten{X}_\text{rec}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
NMSE is calculated as
\begin{equation}
	\nmse\left(\ten{X}, \ten{X}_\text{rec}^{(B)}\right) =
	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{X}(n)\right\rVert_F^2}
\end{equation}
\begin{figure*}[t]
	\input{figures/bttda/blocks.pgf}
	\caption{Normalized Mean Square Error (NMSE) (left), and difference in
    classification score for the training and validation for the ERP datasets
    (top row) and the MI dataset (bottom row)	of the greedy model selection
    procedure (right), as a function of the number of BTTDA blocks $b$.
		While NMSE monotonically decreases for the evaluated datasets, better class
		separation will be achieved, but eventually overfitting occurs and validation
		and test scores will drop, or plateau due to feature selection.
	}
	\label{fig:blocks}
\end{figure*}
\todo{format figure}

@@

Additionally, the forward modeling step inherent to the BTTDA results
in an interpretable model, since we can use the activation patterns or the
forward projection to inspect the neural patterns corresponding to the
relevant discriminatory information at each block~\cite{Haufe2014}.
Figure~\ref{fig:forward} shows the activation patterns
of two blocks obtained from the BNCI2014-008 dataset as well as the forward
projection of the difference between the averages of the mean latent tensor per
class (\emph{contrasts}) after forward projection.
While the weights of the backward projection are
uninterpretable~\cite{Haufe2014},
the activation patterns and contrasts after forward projection clearly show
that ERP components can be recognized and separated into different
BTTDA blocks.
\begin{figure*}[t]
	\includegraphics[width=\linewidth]{figures/bttda/forward_block-0.png}
	\includegraphics[width=\linewidth]{figures/bttda/forward_block-1.png}
	\caption{Spatial (left two columns) and temporal (middle column) activation patterns and
		condition contrasts (right column) obtained after forward projection of the latent
		features for 2 blocks (row 1 and row 2) of rank $(2,2)$ of BTTDA
		fit on the full dataset BNCI2014-008.
		The separate blocks approximately model different ERP
		components.}
	\label{fig:forward}
\end{figure*}
Given informed or correctly tuned hyperparameters, this method could be used to
e.g. separate ERP components or neural processes based on the task-related
information in the class labels.

>>>>

\subsection{Higher Order Discriminant Analysis}
Higher Order Discriminant Analysis (HODA)~\cite{Phan2010} is a
supervised tensor-based feature extraction technique.
For a set of $N$ tensors of order $K$
$\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
		D_K}\right\}_n^N$, HODA finds projection matrices $\mat{U_k}$ for each mode $k$
that project a given $\ten{X}$ to a latent tensor
$\ten{G}\in\mathbb{R}^{r_1\times r_2\times\cdots\times r_K}$, usually with lower
dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
tensor-matrix mode products
\begin{equation}
	\ten{G}  = \ten{X}\mmpr{\mat{U}}
	\label{eq:HODA-backward}
\end{equation}
visualized in Figure~\ref{fig:HODA-backward}.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_backward.tikz.tex}
  \caption[The \ac{hoda} backward projection.]{%
    A visualization of the multilinear projection obtained by Higher Order
    Discriminant Analysis (HODA) applied to a third-order tensor
    sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
		HODA finds projection matrices $\mat{U}_k$ such that maximal
		discriminability between classes is achieved in the projected latent tensors
		$\ten{G}$ with reduced dimensionality $(r_1,r_2,r_3)$.}
	\label{fig:HODA-backward}
\end{figure}
Analogous to the HOSVD, HODA is a dimensionality
reduction decomposition that results in a dense latent tensor $\ten{G}$, and
imposes an orthogonality constraint on $\mat{U}_k$ to ensure uniqueness.
However, while for the HOSVD decomposition the projection matrices
are chosen to minimize the reconstruction error, the projection matrices
$\mat{U}_k$ of HODA are optimized for maximal discriminability of tensors
$\ten{G}(n)$ belonging to classes with labels $c_n$.
This is a desirable property in a classification setting where samples are
high-dimensional tensors.

HODA optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent
tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
The goal is now to find the optimal projection matrices\todo{You should say
somewhere that the shape (r1,r2,...rK) is fixed a priori, no?}
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which can be solved by the iterative algorithm in Algorithm~\ref{alg:HODA}.
\begin{algorithm}
  \caption[The \ac{hoda} backward solution.]{The \ac{hoda} backward solution.}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per mode Singular Value Decomposition (SVD),
or as the partial HOSVD of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
%and solving the generalized eigenvalue problem
%\begin{equation}
%  = \mat{\lambda}\mat{S}_{-k,\text{w}}\mat{U}_k
%\end{equation}
The iterative process halts after a fixed amount or iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.

@@

The HODA projection in Equation~\ref{eq:hoda-backward}\todo{missing} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model is given by
\begin{equation}
	\ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$
and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA. By calculating activation
		patterns $A_k$, the original tensor $\ten{X}$ can approximately be
		reconstructed from projected latent tensor $\ten{G}$. $A_k$ are chosen such
		that the variability captured in the latent tensor is maximally expressed in
		the reconstructed tensor and not in the error term.}
	\label{fig:HODA-forward}
\end{figure}
\todo{not referred to in text}
\todo{I would prefer a figure in which the Ai and G are on the left and
$\hat{X}$on the right. So, rather than using a "almost equal to" sign you could
use a standard equal signal. Finally, in the caption, you could clearly define
the original(X) and reconstructed ($\hat{X}$) tensors as well as the error term.}
\todo{explicitly state why choosing this versus full model or HOPLS}

A forward model should make sure that reconstruction error is minimized and
variation captured in the latent tensor is maximally captured by the forward
projection term $\ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
\todo{not clear ("error is minimized" in the same sentence)}
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares (ALS) algorithm~\cite{Bentbib2022},%~\cite{Comon2009},
iteratively fixing all but one of the activation patterns such that
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed amount of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in Algorithm~\ref{alg:HODA-fw}.
\begin{algorithm}
  \caption[The \ac{hoda} forward solution.]{The \ac{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}
\todo{Reorder figures and add error term}

\subsection{Block-Term Tensor Discriminant Analysis}
After defining the forward model, we can construct our proposed block-term
tensor model.
Assuming the latent tensors $\ten{G}$
obtained by the backward projection of HODA do not achieve perfect
class separation, the error term $\ten{\mathbfcal{E}}$ in
Equation~\ref{eq:HODA-forward} should still contain some discriminative
information, which can be exploited to improve classifier
performance.
\todo{eps1 = X - G1 x {A1} still contains discriminative info, so
it is further projected onto another core tensor yielding
eps2 = eps1 - G2 x
{A2}
etc. and finally you give equation (15)
}
We thus extend the HODA feature extraction scheme with backward and
forward models defined in respectively Equations~\ref{eq:HODA-backward}
and~\ref{eq:HODA-forward} to Block-Term Tensor Discriminant Analysis
(BTTDA).
BTTDA finds multiple discriminative blocks, such that its forward
model adheres to the block-term tensor structure:
\begin{align}
	\ten{X} & = \sum_b^B\ten{G}^{(b)}\mmpr{\mat{A}^{(b)}} + \ten{\mathbfcal{E}}
	\label{eq:BTTDA-forward}
\end{align}
for $B$ extracted latent tensors $\ten{G}^{(b)}$ and residual error term
$\ten{\mathbfcal{E}}$.
Figure~\ref{fig:BTTDA} further illustrates the BTTDA model.
\begin{figure*}[t]
	\centering
	\input{figures/bttda/BTTDA.tikz.tex}
  \caption[A forward model for \ac{bttda}.]{A forward model for Block-Term Tensor Discriminant Analysis
		(BTTDA). BTTDA can extract more features
		than HODA by iteratively finding a latent tensor $\ten{G}^{(b)}$ in a
		deflation scheme.
		The HODA backward projection is first applied. Next, the
		input data is reconstructed via the HODA forward model and the
		difference between the two is found.
		Finally, this process is repeated with this difference as input data, until a
		desired number of blocks $B$ has been found.}
	\label{fig:BTTDA}
\end{figure*}
\todo{Explicitly mention that if restricted to rank one, it is parafac, and to
one block, it is hoda}

Since BTTDA is specified as a forward model, a backward modelling
procedure is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
BTTDA to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in Algorithm~\ref{alg:BTTDA}.
\begin{algorithm}
  \caption{\Ac{bttda}.}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_bttda.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \todo{ref equation}from a residual error term
$\ten{\mathbfcal{E}}^{(b)}$
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward HODA
projection
\begin{equation}
	\ten{\mathbfcal{E}}^{(b+1)} = \ten{\mathbfcal{E}}^{(b)} - \ten{G}^{(b)}
	\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(1)}=\ten{X}$.

@@

Combined, this results in hyperparameters $B$ for the number of blocks, the
threshold value for feature retention, and the ranks of the individual blocks.
While we can reasonably set the feature selection $p$-value threshold to $0.05$,
$B$ and the block ranks must be tuned to select a performant feature extraction
model.
While these hyperparameters can be set through cross-validation, this can be
computationally expensive.
To reduce the computational cost of model selection,
Algorithm~\ref{alg:model-selection} proposes a heuristic model selection
algorithm that leverages cross-validation in a greedy way per block, to
iteratively find the optimal rank for the next block given the ranks of the
previous block.
The area under the receiver operating characteristic curve (ROC-AUC) for
classification of extracted feature vectors after feature selection is used as
cross-validation score.
\begin{algorithm}
  \caption[The greedy model selection procedure.]{Greedy model selection}
	\label{alg:model-selection}
	\input{chapters/bttda/alg_model_selection.tex}
\end{algorithm}
Finally, the series of blocks can be truncated to the point with
highest validation score to determine $B$.

@@

\subsection{Block contribution}
\todo{Discuss number of features}
To analyze the contribution of extra feature blocks extracted by {BTTDA} over
the first one found by HODA we pick n ERP ($K=2$) dataset
(BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
We report within-session ROC-AUC scores for training, validation and test data as a function
of the number of blocks increases, Figure~\ref{fig:blocks}.
Training and validation folds were taken from the model selection procedure.
Additionally, the Normalized Mean Squared Error (NMSE) is reported for the
reconstructed from the truncated BTTDA decomposition
$\textstyle{\ten{X}_\text{rec}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
NMSE is calculated as
\begin{equation}
	\nmse\left(\ten{X}, \ten{X}_\text{rec}^{(B)}\right) =
	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{X}(n)\right\rVert_F^2}
\end{equation}
\begin{figure*}[t]
	\input{figures/bttda/blocks.pgf}
  \caption[Analysis of NMSE and classification score per block.]{%
    Normalized Mean Square Error (NMSE) (left), and difference in
    classification score for the training and validation for the ERP datasets
    (top row) and the MI dataset (bottom row)	of the greedy model selection
    procedure (right), as a function of the number of BTTDA blocks $b$.
		While NMSE monotonically decreases for the evaluated datasets, better class
		separation will be achieved, but eventually overfitting occurs and validation
		and test scores will drop, or plateau due to feature selection.
	}
	\label{fig:blocks}
\end{figure*}
\todo{format figure}

@@

Additionally, the forward modeling step inherent to the BTTDA results
in an interpretable model, since we can use the activation patterns or the
forward projection to inspect the neural patterns corresponding to the
relevant discriminatory information at each block~\cite{Haufe2014}.
Figure~\ref{fig:forward} shows the activation patterns
of two blocks obtained from the BNCI2014-008 dataset as well as the forward
projection of the difference between the averages of the mean latent tensor per
class (\emph{contrasts}) after forward projection.
While the weights of the backward projection are
uninterpretable~\cite{Haufe2014},
the activation patterns and contrasts after forward projection clearly show
that ERP components can be recognized and separated into different
BTTDA blocks.
\begin{figure*}[t]
	\includegraphics[width=\linewidth]{figures/bttda/forward_block-0.png}
	\includegraphics[width=\linewidth]{figures/bttda/forward_block-1.png}
  \caption[Extracted \ac{bttda} activation patterns.]{%
    Spatial (left two columns) and temporal (middle column) activation patterns and
		condition contrasts (right column) obtained after forward projection of the latent
		features for 2 blocks (row 1 and row 2) of rank $(2,2)$ of BTTDA
		fit on the full dataset BNCI2014-008.
		The separate blocks approximately model different ERP
		components.}
	\label{fig:forward}
\end{figure*}
Given informed or correctly tuned hyperparameters, this method could be used to
e.g. separate ERP components or neural processes based on the task-related
information in the class labels.

### 2024-10-02 chapters/conclusion.tex
Comments Marc

\section{Current \& future work}
multicomponent alignment approach
hankel tensors to cope with jitter
select best classifier using eye tracking
expand on patient experiemnts

>>>>

\section{Current \& future work}
kronecker-sum lda
multicomponent alignment approach
hankel tensors to cope with jitter
select best classifier using eye tracking
expand on patient experiemnts
